# (PART\*) R crash course{-}


# Setting R

R is a language and environment for statistical computing and graphics, which can be extended through packages. R was built as an open-source project, meaning every package and software is free! You can visit their [site](https://www.r-project.org/about.html) if you are interested in more technical aspects. To start using R, you need to install two complementary things:

1. **R:** is the raw software that will run the code

2. **R Studio:** is an Integrated  Development Software (IDE), a computer application that provides comprehensive services to make it easier for the developer or programmer to develop software. 

Both installations are designed to be simple and accessible:

1. **R**, [The Comprehensive R Archive Network (CRAN)](https://cran.r-project.org/), can be downloaded for any platform. After clicking the link, choose the appropriate (Linux, (Mac) OS X, or Windows) installation file for R. Download and install R on your computer.

2. **R studio** provides the previous step and its installation in the following [link](https://www.rstudio.com/products/rstudio/download/#download).

A nice overview of R Studio interface and windows can be found at [RStudio Layout](https://rbasics.netlify.app/3-rstudiobasics.html#working-in-rstudio-server). The reference works through an R Studio server; since our recommendation and approach is to download R studio, you don't have to worry about logging in.


# R Basics

Lucky for you, ITAM has an available [GitHub Student Developer Pack](https://education.github.com/pack) for every active student! GitHub Education gives students hands-on experience with free access to various developer tools from GitHub partners. 

We recommend activating your free trial for [DataCamp](https://www.datacamp.com/); it provides R, Python, SQL and more programming courses. I will give you **one BIG cookie** for every **statement of accomplishment** you achieve during this semester of DataCamp.

If you don't want to cover the basics in detail, try these small tutorials to familiarize yourself with the R syntax and RStudio:

1. [RStudio and Working Environment](https://rbasics.netlify.app/3-rstudiobasics.html#working-in-rstudio-server)
2. [R syntax, basic objects and operations](http://r-statistics.co/R-Tutorial.html)

Further, we will cover some essential aspects just as the bare minimum requirements for understanding these notes to some extent.

## Installing Packages

As an open-source project, R has a **universe of add-ons**; additional content is included as a package. For **every idea or problem** you may encounter, **someone has already worked on it** and published it somewhere.

1. To install a package, you need to type `install.packages("package_name")` into the R console or an R script. Press the `enter` key if you are in the R console or press the `shift + enter` key if you are in an R script.


```{r install, eval = FALSE, echo = TRUE}

install.packages("estimatr") # For robust standard errors

```
2. To load a package, you must type `library(package_name)`. Press the `enter` key if you are in the R console or press the `shift + enter` key if you are in an R script.

```{r load, eval = FALSE, echo = TRUE}

library(estimatr) # Loading the library

```

You only need to **install a package once**, but you **must load a package every time** you close and restart R; R needs to know which resources you plan to work with on each session.

# Data wrangling

```{r, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 

```

Data manipulation is surprisingly easy with R. You only need the `dplyr` library included in the `tidyverse` package and know a few words.

## Data Set

This section will work with the Iris data set inherited as default R.  Iris was introduced by Ronald Fisher in his 1936 paper __The use of multiple measurements in taxonomic problems__, contains **three plant species** (setosa, virginica, versicolor) and **four features** measured for each sample.


```{r intro, echo = TRUE}

# The datasets package needs to be loaded to access our data 
# For a full list of these datasets, type library(help = "datasets")
library(datasets)
data(iris)
```

```{r intro_d, echo = FALSE}

library(kableExtra)

head(iris) %>% 
  kbl() %>%
  kable_paper("hover", full_width = F)
  
```


## Dplyr's Wonder

`dplyr` package allows **easy manipulation** through **"verbs"** with a novel pipeline feature. The pipelines shine when you need to arrange many features, which is usually the case.

- `dplyr` verbs are the name of the functions to arrange the data; these include filter, mutate, arrange, group_by and more.

- **The pipeline** `%>%`  means _take the name of the previous object and put it in the first argument of the next function_. The pipelines shine when you need to arrange many features, which is usually the case.

```{r , eval = FALSE}

library(dplyr)

# 1. Filter without dplyr
example <- filter(iris, Sepal.Width == 3.5)

# 2. Filter with dplyr
example <- iris %>% 
           filter( Sepal.Width == 3.5)

```

Here we will cover 3 basic verbs you might need in your journey; you can consult the [most useful manipulation tips](https://www.maximaformacion.es/blog-ciencia-datos/top-10-manipulacion-de-datos-con-dplyr/) here.


#### 1. Add or change variables {-}

```{r mutate, echo = TRUE}

library(dplyr)

# 1. The mutate verb

example <- iris %>% 
           # A. You can add a complete new variable
           mutate( new_var = 3.3+6/5) %>% 
  
           # B. You can change current variables
           mutate(  Sepal.Width = new_var + Sepal.Width)

```

```{r mut_ex, echo = FALSE}

head(example) %>% 
  kbl() %>%
  kable_paper("hover", full_width = F) %>%
  column_spec(2, background = "#d8d9fb") %>% 
  column_spec(6, background = "#d8d9fb")
  
```


#### 2. Filter the data set {-}

```{r filter, echo = TRUE}

# 2. The filter verb

example <- iris %>% 
           # A. You can filter with specific logical criteria
           filter( Sepal.Length > 5) %>% 
  
           # B. You can add multiple conditions
           filter( Sepal.Length > 5, Sepal.Width < 3)

```

```{r filt_ex, echo = FALSE}

head(example) %>% 
  kbl() %>%
  kable_paper("hover", full_width = F)%>%
  column_spec(1, background = "#d8d9fb") %>% 
  column_spec(2, background = "#d8d9fb")
  
```


#### 3. Select variables of interest{-}

```{r select, echo = TRUE}

# 3. The select verb

example <- iris %>% 
           # A. You can select existing variables
           select( Sepal.Length, Sepal.Width, Species) %>% 
  
           # B. You can also remove just one
           select( -Species)

```

```{r select_ex, echo = FALSE}

head(example) %>% 
  kbl() %>%
  kable_paper("hover", full_width = F)
  
```



